Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2014-09-07 pkg/paper/Paper/paper.Rnw
preparation for new release accompanying EEE paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@131 b679b2f7-6539-45d0-a605-5ddcb976cde9

\documentclass[nojss]{jss}
\usepackage{amsfonts,amstext,amsmath,thumbpdf}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

@@

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Faculty of Economics and Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstra{\ss}e 15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
  Telephone: +43/12/507-7103 \\
  Fax: +43/512/507-2851 \\
}

@@

The null hypothesis of parameter stability (\ref{eq:h0}) can be assessed using the empirical
scores $s(y_i | \hat \phi)$ as measures of model deviation \citep{Zeileis+Hornik:2007} from the
model fit under $H_0$. Systematic deviations over time from the full sample estimates $\hat \phi$
can then be captured by cumulative sums of the empirical scores in an empirical fluctuation
process $\mathit{efp}(t)$:
%
\begin{equation} \label{eq:efp}
\mathit{efp}(t) ~=~ \hat V^{-1/2} n^{-1/2} \sum_{i = 1}^{\lfloor nt \rfloor}
  s(y_i ~|~ \hat \phi) \quad (0 \leq t \leq 1),
\end{equation}
%
where $\hat V$ is some consistent estimator of the variance of the scores. Below,
we use the outer product of gradients $s(y_i | \hat \phi)$ to estimate $\hat V$.
Alternatively, HAC estimators could be used leading to the same qualitative
results for the data under investigation.

@@

\begin{figure}[t!]
<<slovenia-plots1, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
plot(x$data, main = "", ylab = "Monthly inflation rate (in percent)", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), lwd = 2)
if(x$nbreaks > 0) lines(confint(x$breakpoints, breaks = x$nbreaks))

plot(x$efp, functional = supLM(0.1), main = "", ylab = "LM statistics", ylim = c(0, 40), boundary = FALSE)
abline(h = supLM(0.1)$computeCritval(0.05, 3), col = "darkgray", lwd = 1.5)
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{\label{fig:slovenia1} Inflation rate series for Slovenia
with breakpoint estimate (\Sexpr{breakdates(x$breakpoints)}), associated confidence interval, and fitted mean
from generalized logistic distribution (left). Corresponding sup$\mathit{LM}$ test (right)
with sequence of $\mathit{LM}$~statistics,
critical value at 5\% level (horizontal gray line), and estimated breakpoint (vertical dashed line).}

@@

As there is evidence for at least one break, the LWZ criterion is employed for $B = 1, 2, \dots$
breakpoints with a minimal segment size of two years per segment. It assumes its minimum for $B = 1$
associated with the breakpoint \Sexpr{breakdates(x$breakpoints, breaks = x$nbreaks)}, depicted
by a vertical dashed line in both panels of Figure~\ref{fig:slovenia1}. The resulting segmented
fitted mean (horizontal blue lines in the left panel of Figure~\ref{fig:slovenia1}) shows that
before the break, Slovenia experienced very high inflation rates but was successful in reducing
inflation to a much lower level afterwards.

@@

Austria, Finland, and France adjusted to the requirements of the EMU during the
first phase of convergence. In these countries, the result of the adjustment process
was a lower inflation rate and an overall decrease in the variance.
Additionally, if a HAC correction is used, the change for France would be nonsignificant,
suggesting that France may also belong to the group of countries which did not change at all.

@@

The next group consists of two countries that, after a very short participation in the
ERM, decided not to be part of it and hence also not to participate in the EMU:
the United Kingdom and Sweden. Both show clear breaks in the early 1990s which can be
traced back to economic crises: the currency crisis of the United Kingdom cumulating
into the ``Black Wednesday'' in 1992 and the banking crisis in Sweden during the early
1990s. With regard to the last segment for both countries, the Ljung-Box test suggests that 
the autocorrelation, albeit low around 0.16, is significant. However, both the standard
and the HAC-corrected sup$\mathit{LM}$ lead to highly significant evidence for structural
breaks.

@@

Finally, the inflation rates for the whole euro area\footnote{This is the
euro area 17~countries series as provided by the \citet{oecd},
starting in \Sexpr{start(eu) - 1/12}. It is calculated as a weighted average of the HICPs
of now seventeen euro countries.} are considered and
three different regimes are found: (1)~until 1999 with both a low mean and
variance, (2)~the third and final stage of monetary integration, accompanied
by a significant increase in the mean, (3)~a period encompassing the
overheating of the economy and its culmination in the financial crisis, showing a
large increase in variance and a lower mean inflation.

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\centering
<<glmean-plot, fig=TRUE, height=5, width=7.5>>=
gray_red <- rgb(c(0.2, 0.8), c(0.2, 0.0), c(0.2, 0.0), alpha = 0.3)
gray_red <- rgb(c(0.2, 0.0), c(0.2, 0.0), c(0.2, 0.0), alpha = c(0.3, 0.5)) ## actually gray-darkgray now
plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1],
  xlab = "Time", ylab = "Monthly inflation rates")
for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{\label{fig:glmean} Observed monthly inflation rates (light gray) and estimated means from
12-month rolling GL fits (dark red) over time for all countries.}
\end{figure}
\setkeys{Gin}{width=\textwidth}

@@

Our results were obtained using 
\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \citep{R} with the packages
\pkg{glogis}~\Sexpr{packageDescription("glogis")["Version"]},
\pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \citep{Zeileis+Leisch+Hornik:2002}, and
\pkg{fxregime}~\Sexpr{packageDescription("fxregime")["Version"]} \citep{Zeileis+Shah+Patnaik:2010},
all of which are available under the General Public License (GPL) from the
Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}). Package \pkg{strucchange} provides
various techniques for inference in structural-change settings which are complemented
with additional maximum-likelihood-based methods by \pkg{fxregime}. Inference for the
generalized logistic model (that can be combined with \pkg{strucchange}/\pkg{fxregime})
is implemented in the new package \pkg{glogis} that also contains the data used, both
the raw monthly price indexes \citep[as provided by the][]{oecd} and the seasonally adjusted
monthly inflation rates. The seasonal adjustment was performed in \pkg{X-12-ARIMA}~0.3
(\citealp{X-12-ARIMA}, see also \citealp{Findley+Monsell+Bell:1998}) through the interface
provided by \proglang{gretl}~1.9.5 (\citealp{gretl}, see also \citealp{Smith+Mixon:2006} for a review).

>>>>

\documentclass[nojss,noheadings]{jss}
\usepackage{amsfonts,amstext,amsmath,thumbpdf}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

@@

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Faculty of Economics and Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstra{\ss}e 15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Achim.Zeileis@R-project.org} \\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}

%% publication info and copyright notice
\usepackage{fancyhdr}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0pt}
\pagestyle{fancy}
\thispagestyle{plain}
\fancyhf{}
\fancyhead[RO,LE]{\thepage}
\fancyhead[CO]{\textit{\normalsize Thomas Windberger, Achim Zeileis}}
\fancyhead[CE]{\textit{\normalsize Structural Breaks in Inflation Dynamics within the European Monetary Union}}
\fancyfoot[LO,LE]{\small Copyright {\copyright} 2014 M.E. Sharpe, Inc.}
\fancypagestyle{plain}{
\fancyhf{}
\fancyfoot[LO,LE]{\small This is a preprint of an article published in \textit{Eastern European Economics}, \textbf{52}(3), 67--89.
  \doi{10.2753/EEE0012-8775520304}.
  Copyright~{\copyright} 2014 M.E. Sharpe, Inc.
  }
}

@@

The null hypothesis of parameter stability (\ref{eq:h0}) can be assessed using the empirical
scores $s(y_i | \hat \phi)$ as measures of model deviation from the model fit under $H_0$
\citep[as in][]{Zeileis+Hornik:2007}. Systematic deviations over time from the full sample estimates $\hat \phi$
can then be captured by cumulative sums of the empirical scores in an empirical fluctuation
process $\mathit{efp}(t)$:
%
\begin{equation} \label{eq:efp}
\mathit{efp}(t) ~=~ \hat V^{-1/2} n^{-1/2} \sum_{i = 1}^{\lfloor nt \rfloor}
  s(y_i ~|~ \hat \phi) \quad (0 \leq t \leq 1),
\end{equation}
%
where $\hat V$ is some consistent estimator of the variance of the scores. Below,
we use the outer product of gradients $s(y_i | \hat \phi)$ to estimate $\hat V$.
Alternatively, HAC estimators could be used leading to the same qualitative
results for the data under investigation.

@@

\begin{figure}[t!]
<<slovenia-plots1, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
plot(x$data, main = "", ylab = "Monthly inflation rate (in percent)", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), lwd = 2)
if(x$nbreaks > 0) lines(confint(x$breakpoints, breaks = x$nbreaks), col = "darkgray")

plot(x$efp, functional = supLM(0.1), main = "", ylab = "LM statistics", ylim = c(0, 40), boundary = FALSE)
abline(h = supLM(0.1)$computeCritval(0.05, 3), col = "darkgray", lwd = 1.5)
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{\label{fig:slovenia1} 
  Inflation rate series for Slovenia with breakpoint estimate (\Sexpr{breakdates(x$breakpoints)})
  and associated sequence of LM statistics
  Left: Inflation rate series with breakpoint estimate (vertical dashed line),
  associated confidence interval (gray, at the bottom), and fitted mean from generalized
  logistic distribution (bold horizontal lines). Right: Corresponding sup$\mathit{LM}$ test
  with sequence of LM statistics, critical value at 5\% level (horizontal gray
  line), and estimated breakpoint (vertical dashed line).
}

@@

As there is evidence for at least one break, the LWZ criterion is employed for $B = 1, 2, \dots$
breakpoints with a minimal segment size of two years per segment. It assumes its minimum for $B = 1$
associated with the breakpoint \Sexpr{breakdates(x$breakpoints, breaks = x$nbreaks)}, depicted
by a vertical dashed line in both panels of Figure~\ref{fig:slovenia1}. The resulting segmented
fitted mean (horizontal bold lines in the left panel of Figure~\ref{fig:slovenia1}) shows that
before the break, Slovenia experienced very high inflation rates but was successful in reducing
inflation to a much lower level afterwards.

@@

The second group of countries, Austria, Finland, and France, adjusted to the requirements of the EMU during the
first phase of convergence. In these countries, the result of the adjustment process
was a lower inflation rate and an overall decrease in the variance.
Additionally, if a HAC correction is used, the change for France would be nonsignificant,
suggesting that France may also belong to the group of countries which did not change at all.

@@

The fifth group consists of two countries that, after a very short participation in the
ERM, decided not to be part of it and hence also not to participate in the EMU:
the United Kingdom and Sweden. Both show clear breaks in the early 1990s which can be
traced back to economic crises: the currency crisis of the United Kingdom cumulating
in ``Black Wednesday'' in 1992 and the banking crisis in Sweden during the early
1990s. With regard to the last segment for both countries, the Ljung-Box test suggests that 
the autocorrelation, albeit low around 0.16, is significant. However, both the standard
and the HAC-corrected sup$\mathit{LM}$ lead to highly significant evidence for structural
breaks.

@@

Finally, the inflation rates for the whole euro area\footnote{This is the
EU17 series as provided by the \citet{oecd},
starting in \Sexpr{start(eu) - 1/12}. It is calculated as a weighted average of the HICPs
of the then seventeen euro countries.} are considered and
three different regimes are found: (1)~until 1999 with both a low mean and
variance, (2)~the third and final stage of monetary integration, accompanied
by a significant increase in the mean, (3)~a period encompassing the
overheating of the economy and its culmination in the financial crisis, showing a
large increase in variance and a lower mean inflation.

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\centering
<<glmean-plot, fig=TRUE, height=5, width=7.5>>=
gray_red <- rgb(c(0.2, 0.8), c(0.2, 0.0), c(0.2, 0.0), alpha = 0.3)
gray_red <- rgb(c(0.2, 0.0), c(0.2, 0.0), c(0.2, 0.0), alpha = c(0.2, 0.5)) ## actually gray-darkgray now
plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1],
  xlab = "Time", ylab = "Monthly inflation rates")
for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{\label{fig:glmean} Observed monthly inflation rates (light gray) and estimated means from
12-month rolling GL fits (dark red) over time for all countries.}
\end{figure}
\setkeys{Gin}{width=\textwidth}

@@

Our results were obtained using 
\proglang{R}~3.0.0 %\Sexpr{paste(R.Version()[6:7], collapse = ".")}
\citep{R} with the packages
\pkg{glogis}~0.1-0, %\Sexpr{packageDescription("glogis")["Version"]},
\pkg{strucchange}~1.4-7 %\Sexpr{packageDescription("strucchange")["Version"]}
\citep{Zeileis+Leisch+Hornik:2002}, and
\pkg{fxregime}~1.0-2 %\Sexpr{packageDescription("fxregime")["Version"]}
\citep{Zeileis+Shah+Patnaik:2010},
all of which are available under the General Public License (GPL) from the
Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}). Package \pkg{strucchange} provides
various techniques for inference in structural-change settings which are complemented
with additional maximum-likelihood-based methods by \pkg{fxregime}. Inference for the
generalized logistic model (that can be combined with \pkg{strucchange}/\pkg{fxregime})
is implemented in the new package \pkg{glogis} that also contains the data used, both
the raw monthly price indexes \citep[as provided by the][]{oecd} and the seasonally adjusted
monthly inflation rates. The seasonal adjustment was performed in \pkg{X-12-ARIMA}~0.3
(\citealp{X-12-ARIMA}, see also \citealp{Findley+Monsell+Bell:1998}) through the interface
provided by \proglang{gretl}~1.9.5 (\citealp{gretl}, see also \citealp{Smith+Mixon:2006} for a review).
